---
title: A couple of case_when() tricks
author: Alan Yeung
date: '2023-10-09'
slug: couple-of-casewhen-tricks
categories:
  - R
tags:
  - tidyverse
  - data manipulation
image:
  caption: ''
  focal_point: ''
draft: no
toc: false
---

## Combining case_when() and across()
If you want to use `case_when()` and `across()` different variables, then here is an example that can do this with the help of the `get()` and `cur_column()` functions.

```{r casewhen-across, message=FALSE}
library(tidyverse)

iris_df <- as_tibble(iris) %>% 
  mutate(flag_Petal.Length = as.integer(Petal.Length > 1.5),
         flag_Petal.Width = as.integer(Petal.Width > 0.2))

iris_df %>% 
  mutate(across(c(Petal.Length, Petal.Width),
                ~case_when(
                  get(glue::glue("flag_{cur_column()}")) == 1 ~ NA_real_,
                  TRUE ~ .x
                ))) %>% 
    select(contains("Petal"))
```

Is there any way to use this with a function from the `rlang` packages instead of `get()`? It's a little beyond my current understanding of tidy evaluation but let me know in the comments if you know please.


## Combining case_when() with if_any() and if_all()
The `if_any()` and `if_all()` functions can be used to save typing lots of variables (as these allow the use of `tidyselect` helpers) within `case_when()`.

```{r casewhen-if}
df1 <- tibble(var1 = c(1, 0, 1, NA),
              var2 = c(1, 0, 0, NA))

df1 %>% 
  mutate(category = case_when(
    if_any(c(var1, var2), ~.x > 0) ~ 1, 
    if_all(c(var1, var2), is.na) ~ NA_real_,
    TRUE ~ 0
  ))
```
