---
title: Grouped Sequences in dplyr Part 2
author: Alan Yeung
date: '2023-11-07'
slug: grouped-sequences-in-dplyr-part-2
categories:
  - R
tags:
  - data manipulation
  - tidyverse
image:
  caption: ''
  focal_point: ''
draft: no
toc: false
---

I just wrote a post about [grouped sequences in dplyr](../2023-11-05-grouped-sequences-in-dplyr) and following that, I've been made aware of another couple of solutions to this problem (credit [John Mackintosh](https://johnmackintosh.net/){target="_blank"}). The solution involves using the `consecutive_id()` function, available in dplyr since v1.1.0. In the help page for this function, it's mentioned that it was inspired by `rleid()` function from the `data.table` package. These functions work similarly to the `rle()` function I used last time (in what I called 'the complicated solution') but provide neater outputs. I'll try them both out here.

```{r dummy-data, message=FALSE}
library(tidyverse)

sports_df <- tibble::tribble(
  ~team,    ~date,        ~home_or_away, ~num_matches_away,
  "Team A", "07/10/2022", "H",           0L,
  "Team A", "14/10/2022", "A",           1L,
  "Team A", "21/10/2022", "A",           2L,
  "Team A", "28/10/2022", "H",           0L,
  "Team A", "04/11/2022", "A",           1L,
  "Team A", "11/11/2022", "A",           2L,
  "Team A", "18/11/2022", "A",           3L,
  "Team A", "25/11/2022", "H",           0L,
  "Team A", "02/12/2022", "H",           0L,
  "Team A", "09/12/2022", "H",           0L,
  "Team B", "07/10/2022", "H",           0L,
  "Team B", "14/10/2022", "A",           1L,
  "Team B", "21/10/2022", "A",           2L,
  "Team B", "28/10/2022", "A",           3L,
  "Team B", "04/11/2022", "H",           0L,
  "Team B", "11/11/2022", "A",           1L,
  "Team B", "18/11/2022", "A",           2L,
  "Team B", "25/11/2022", "H",           0L,
  "Team B", "02/12/2022", "H",           0L,
  "Team B", "09/12/2022", "A",           1L
) %>%      
  mutate(date = as.Date(date, "%d/%m/%Y")) %>% 
  arrange(team, date)
```

## dplyr::consecutive_id()

```{r consec-id}
sports_df2 <- sports_df %>% 
  group_by(team) %>% 
  mutate(consec_id = consecutive_id(home_or_away)) %>% 
  group_by(team, consec_id) %>% 
  mutate(days_on_road = 1:n(),
         days_on_road = if_else(home_or_away == "H", 0L, days_on_road)) %>% 
  ungroup()

sports_df2
identical(sports_df2$days_on_road, sports_df$num_matches_away)
```

## data.table::rleid()

```{r rleid}
sports_df3 <- sports_df %>% 
  group_by(team) %>% 
  mutate(rleid_id = data.table::rleid(home_or_away)) %>% 
  group_by(team, rleid_id) %>% 
  mutate(days_on_road = 1:n(),
         days_on_road = if_else(home_or_away == "H", 0L, days_on_road)) %>% 
  ungroup()

sports_df3
identical(sports_df3$days_on_road, sports_df$num_matches_away)
```

We now have 4 solutions to this problem - it's often the case that there are many ways to do the same thing in R! It can be extra effort to remember different solutions but they work in different ways and can be applied to different problems so I do think it's worthwhile trying these out. Which solution is your favourite?
