---
title: Summarising Dates with Missing Values
author: Alan Yeung
date: '2023-10-02'
slug: summarising-dates-with-missing-values
categories:
  - R
tags:
  - tidyverse
  - dates
draft: no
image:
  caption: ''
  focal_point: ''
toc: false
---

This blog post is just a note that when you try to do a grouped summary of a date variable but some groups have **all** missing values, it will return `Inf`. This means that the summary will not show up as an `NA` and this can cause issues in analysis if you are not careful.

```{r setup-example, eval=FALSE}
library(tidyverse)

df <- tibble::tribble(
    ~id,          ~dt,
    1L, "01/01/2001",
    1L,           NA,
    2L,           NA,
    2L,           NA
) %>% 
    mutate(dt = dmy(dt))

z1 <- df %>% 
    group_by(id) %>% 
    summarise(dt_min = min(dt, na.rm = TRUE),
              .groups = "drop")

z1
# A tibble: 2 × 2
#      id dt_min    
#   <int> <date>    
# 1     1 2001-01-01
# 2     2 Inf

sum(is.na(z1$dt_min))
# [1] 0
```

There are a couple of ways around this. Firstly you can use an `if()` statement.

```{r if, eval=FALSE}
z2 <- df %>% 
    group_by(id) %>% 
    summarise(dt_min = if (all(is.na(dt))) NA_Date_ else min(dt, na.rm = TRUE),
              .groups = "drop")

z2
# A tibble: 2 × 2
#      id dt_min    
#   <int> <date>    
# 1     1 2001-01-01
# 2     2 NA  
sum(is.na(z2$dt_min))
# [1] 1
```

Or you can summary functions from the [`hablar`](https://cran.r-project.org/web/packages/hablar/vignettes/hablar.html) package.

```{r hablar, eval=FALSE}
z3 <- df %>% 
    group_by(id) %>% 
    summarise(dt_min = hablar::min_(dt),
              .groups = "drop")

z3
# A tibble: 2 × 2
#      id dt_min    
#   <int> <date>    
# 1     1 2001-01-01
# 2     2 NA  
sum(is.na(z3$dt_min))
# [1] 1
```

Is there a reason why R decides to return `Inf` when summarising dates? Are there any other solutions to summarising date variables that contain missing values? Leave me a comment if you know thanks.
