---
title: Grouped Sequences in dplyr
author: Alan Yeung
date: '2023-11-05'
slug: []
categories:
  - R
tags:
  - data manipulation
  - tidyverse
image:
  caption: ''
  focal_point: ''
draft: no
toc: false
---

For a piece of work I had to calculate the number of matches that a team plays away from home in a row, which we will call `days_on_the_road`. I was not sure how to do this with `dplyr` but it's basically a 'grouped sequence'. For this post, I've created some dummy data to illustrate this idea. The `num_matches_away` variable is what we want to mimic using some data manipulation.

```{r dummy-date, message=FALSE}
library(tidyverse)

sports_df <- tibble::tribble(
  ~team,    ~date,        ~home_or_away, ~num_matches_away,
  "Team A", "07/10/2022", "H",           0L,
  "Team A", "14/10/2022", "A",           1L,
  "Team A", "21/10/2022", "A",           2L,
  "Team A", "28/10/2022", "H",           0L,
  "Team A", "04/11/2022", "A",           1L,
  "Team A", "11/11/2022", "A",           2L,
  "Team A", "18/11/2022", "A",           3L,
  "Team A", "25/11/2022", "H",           0L,
  "Team A", "02/12/2022", "H",           0L,
  "Team A", "09/12/2022", "H",           0L,
  "Team B", "07/10/2022", "H",           0L,
  "Team B", "14/10/2022", "A",           1L,
  "Team B", "21/10/2022", "A",           2L,
  "Team B", "28/10/2022", "A",           3L,
  "Team B", "04/11/2022", "H",           0L,
  "Team B", "11/11/2022", "A",           1L,
  "Team B", "18/11/2022", "A",           2L,
  "Team B", "25/11/2022", "H",           0L,
  "Team B", "02/12/2022", "H",           0L,
  "Team B", "09/12/2022", "A",           1L
) %>%      
  mutate(date = as.Date(date, "%d/%m/%Y")) %>% 
  arrange(team, date)

sports_df
```

I firstly came up with a complicated solution for this using a combination of `slider::slide()` and `rle()` (run length encoding).

```{r slide-solution}
sports_df2 <- sports_df %>% 
  group_by(team) %>% 
  mutate(days_on_road = unlist(slider::slide(rle(home_or_away)$length, ~1:.x)),
         days_on_road = if_else(home_or_away == "H", 0L, days_on_road)) %>% 
  ungroup()

identical(sports_df2$days_on_road, sports_df$num_matches_away)
```

These are some pretty cool functions and it was nice to know `rle()` existed but I was never really happy with this solution as it seemed overly complex and it's difficult to understand what the code is doing by simply reading it. So I asked a colleague to try to solve this problem and they came up with a better solution which I'm grateful for! It involves using a combination of `group_by()` and `seq_len()` which is a whole lot simpler to understand in my opinion.

```{r grouped-seq}
sports_df3 <- sports_df2 %>%
  mutate(away = cumsum(home_or_away == "H")) %>%
  group_by(team, home_or_away, away) %>%
  mutate(days_on_road = seq_len(n())) %>%
  ungroup() %>%
  mutate(days_on_road = if_else(home_or_away == "H", 0L, days_on_road))

identical(sports_df3$days_on_road, sports_df$num_matches_away)
```

I hope this is useful for some of you out there coding with R!
